---
title: "Project Name"
subtitle: "Project Tagline"
author: "Najma Osman"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  bookdown::pdf_document2:
toc: FALSE
urlcolour: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(gt)
```

# Project Overview (600 words)

## Early Exploration Results

### What did you find?

```{r}

```

```{r}

```

```{r}

```

### Next Steps

## Key Data Elements/Fields

## Key Questions

### Q1
### Q2
### Q3
### Q4
### Q5


# Data

## Source + Type

I've chosen a [public dataset](https://data.ontario.ca/dataset/school-information-and-student-demographics) from Ontario's Data Catalogue. This dataset is on school information and student demographics in the province of Ontario, and includes the following:

* board information
* school information
* grade 3 and 6 EQAO achievements for reading, writing and mathematics
* grade 9 EQAO academic and applied student achievements
* grade 10 OSSLT student achievement
* student demographic percentages on student parents, special education, first language spoken, and new students to Canada

The data is reported by:
* Board School Identification Database (BSID) and Ontario School Information System (OnSIS),
* OnSIS Preliminary 2018-2019 (Student Population)
* the Education Quality and Accountability Office (EQAO), specifically the 2018-2019 data, 
* and the 2016 census.

It discludes private schools, Education and Community Partnership Programs, as well as summer and night schools.

## Cleaning Requirements (500 words)

Does the data set need cleaning? If so, where, how, and why? Have you done any of the cleanings? Please share a link to the cleaned data as well

# Expected Outcomes (400 words)

What do you expect to learn? We are not asking to predict the results but asking what you will be able to tell a story on? Housing prices? Pandemic control? Marketing results?

# Expected Challenges (400 words)

Getting access to the latest data
Finding an expert on the topic
Finding detailed data
Data consistency
...
